What Can Stack Traces Tell us about of the use Exceptions in Java?

### 2014-07-07 android-stacks.tex
small corrections

\begin{document}
\title{What Can Stack Traces Tell us about Exceptions in Java?}

@@

@@

@@

  \item  It performs a large scale analysis of Java stack traces and how
    they can reveal bad-practices on the use of exceptions in Java.

@@

Techniques for error detection and handling are not  an optional add-on but a
fundamental part of any system~\cite{bruntink2006discovering}. Several
approaches exist for detecting errors and responding to them, such as the return-code 
idiom (widely used in systems written in C), and the exception handling 
mechanism, embedded in many main-stream programming languages.

The main disadvantage of return codes is the fact that the programmer can just
ignore the return code and let the program execute in inconsistent state.

The exception mechanism~\cite{goodenough1975exception} was proposed as a way of
preventing errors from being ignored and was embedded in many mainstream
programming languages ~\cite{garcia2001comparative} since then.

In languages such as CLU, Modula-3, Guide, and Extended Ada the exceptions are 
checked, which means that the 
compiler statically checks if appropriate handlers are provided within the system. 
Accordingly, the programmer must explicitly specify every exception that can be
signaled to support the compile time checking \footnote{In CLU,  if a handler is 
found for a given exception, it is converted in a predened default exception, 
called failure which is the only one which can be implicitly 
propagated in this model.}. This approach is known to be
costly to maintain (Dooren and Steegmans, 2005) since exception updates are
cascading; if a new exception type is added to a method signature then all other
methods directly using this method must either change their signatures to
include the new exception type or handle the new exception.

@@

In Java a method can throw either checked or unchecked exceptions. Java is the only language that allows this hybrid approach, on which the developer can define exceptions of both kinds to be thrown and
handled in a single system. Next section details Java exception handling model.

.......

Although exception handling mechanism is the most used technique to deal with such 
failures - been embedded in several programming languages (e.g. Java, C++, C\#) -  
the exception handling code is often generally poorly understood and the least
tested part of software systems.This behaviour may lead to the problem of uncaught exceptions~\cite{jo2004uncaught}, one of the main causes of application crashes
(e.g.~\cite{miller1997issues,Robil00,shah2010understanding, garcia2007extracting, garcia2001comparative,cabral2007exception,coelho2011unveiling}).

........

\subsection{Exception Handling in Java} \label{sec:extypes}

@@

Figure~\ref{fig:overviewfig} presents an overview of the mixed-methods approach
conducted to answer these questions.

@@

The mixed-methods approach is composed by the following steps:

Step1. Retrieving the Dataset to be Analized. We begun our investigation using the dataset 
provided by the GHTorrent project~\cite{Gousi13}, an off-line mirror of the data 
offered through the Github API.  GHTorrent has been collecting data from Git since 
February 2012. Up to Dec 2013, when we queried GHTorrent, it included 16.836 Java projects 
which overall contained 356.057 issues.\footnote {A Java project is specified in GitHub with a
specific label.} From GHTorrent we retrieved the metadata related to each Java project (e.g., 
repository's names and short descriptions), and the content of every issue and issue comment 
and stored in a relational database.

Step 2. Extracting Exception Information from Issues. In this step, we used the ExceptionMiner tool,
 developed in this work, to extract the exception names and exception stack traces of each
 issue and issue comment - analyzing the information of each project at a time. This initial 
set of analyzed Java projects included libraries, web-based, Android and stand-alone applications.

Step 3. Restricting the Dataset to Android Projects. We select a subset of all Android 
projects available on the dataset to perform a deeper analysis (which combined mined
 information extrated from issues with source-code analysis and manual inspection 
of projects resources described on next steps). This decision was driven by the
fact that most open source Anedroid applications are known for dealing
with several sources of exceptions due to multi-threaded execution and
interaction with input/output resources; moreover they share a common
well-defined platform and a limited set of libraries.
To identify Android projects, we performed a case insensitive search for 
the term \textsf{android} in the repository's names and short descriptions.  The
heuristic filtered 2.542 repositories, from which 589 apps had at least one
issue containing a stack trace. Then we performed further cleanup, inspecting the site of every Android
reporting at least one stack trace, to make sure that they represented real
mobile apps. During this clean up 106 apps were removed because they were either
example projects (i.e., toy projects) or tools to support Android development
(e.g. \textsf{selendroid}, \textsf{roboeletric} --- tools to support the testing of Android apps).
The filtered set consisted of 482 apps, from which approximately 50\% are also
hosted in Google Market Place.

Step 4. Combining Information Mined From Issues with Bytecode-Code Analysis.

For each one of the analyzed projects, we
downloaded the source code and using custom scripts, we extracted their package
hierarchy (recursive list of of Java packages) and their custom exception types
as described next.

Step 5
. Combining Information Mined From Issues with Source-Code Analysis.
For each one of the analyzed projects, we
downloaded the source code and using custom scripts, we extracted their package
hierarchy (recursive list of of Java packages) and their custom exception types
as described next.

Step 5. Combining Information Mined From Issues with Manual Inspection.

To support a deeper investigation of the stack traces of Android apps, every
exception defined on a stack trace was then classified according to its type
(e.g. Error, Exception or RuntimeException) (step 4). The exception classifier
was initialized with all the exceptions defined on the Java SDK version 7, the
exceptions of the Android (Platform and OS) (level 19), and all the exceptions
of the 482 Android projects selected for this study, extracted in the previous
step. Moreover, all the packages associated to exceptions and signalers were
also extracted to support further analysis (step 5). The signaler classifier is
initialized with information about all packages found on Java SDK version 7, and
Android Platform and OS (level 19). Finally, additional meta-data is combined to
each stack trace (step 6) which used together with manual inspections (step 7)
enabled to provide answers to RQ3.

@@

If we perform a cross-project analysis, the java.lang.NullPointerException, we could observe that such 
exception was reported on 1,671 projects (which represents 44,47% of all projects reporting at least one stack trace).

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{tab:typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~\ref{tab:typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. After filtering all the exceptions implicitly signaled by 
JVM (due to programming mistakes) and inspecting the signaler methods of such exceptions (i.e.,  direct an inderect 
instances of java.lang.RuntimeException). We could observe that only 1 method out of 118 inspected signalers
 (i.e., 0,8\%) documented the explicitly thrown runtime exception in a Javadoc comment; and none of these methods
included the runtime exception  (reported on the stack) as part of the exception interface of the method (i.e., using 
throws clause on method signature). This result is aligned with the results of other  study conducted by from 
Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=0.4]{stack-wrappings-hist.pdf}
  \end{center}
  \caption{Number of stack traces wrappings for Java and Android repositories. }
  \label{fig:sizewrapbhists}
\end{figure}

>>>>

\begin{document}
\title{What Can Stack Traces Tell us about of the use Exceptions in Java?}

@@

  \item  A multitude of programming mistakes were reported as the main causes of stack traces:
  41,1\%  (12,625)  of reported stack traces were caused by programming mistakes; and 26,31\% of analyzed stack traces reported a java.lang.NullPointerException as root cause;

   \item 1,671 projects (44,47\%) contained at least one issue reporting a java.lang.NullPointerException as the root cause of a stack trace.

  \item Undocumented runtime exceptions signaled by third party
    code (i.e., libraries and Android platform). This study found 118 distinct methods defined on third party code and  reported as the root signalers of runtime exceptions. Only 1 of such methods (0,8\%) documented the explicitly thrown runtime exception in a Javadoc comment; and none of these methods included the exception as part of the exception interface (i.e., using throws clause).

@@

  \item  Issues reporting uncaught checked exceptions.

@@

%Hence, the contributions of this study are as follows:
%\begin{itemize}
%  \item  It performs a large scale analysis of Java stack traces and how
 %   they can reveal bad-practices on the use of exceptions in Java.
% \item  It introduces ExceptionMiner, a tool developed to support cross-repository stack trace
%    analysis.
%\end{itemize}

@@

Techniques for error detection and handling are not  an optional add-on but a
fundamental part of any system~\cite{bruntink2006discovering}. Several
approaches exist for detecting errors and responding to them, such as the return-code 
idiom (widely used in systems written in C), and the exception handling 
mechanism, embedded in many main-stream programming languages.
The main disadvantage of return codes is the fact that the programmer can just
ignore the return code and let the program execute in inconsistent state. 
The exception mechanism~\cite{goodenough1975exception} was proposed as a way of
preventing errors from being ignored and was embedded in many mainstream
programming languages ~\cite{garcia2001comparative} since then.

In languages such as CLU\footnote{In CLU,  if a handler is 
found for a given exception, it is converted in a predened default exception, 
called failure which is the only one which can be implicitly 
propagated in this model [ref-clue].}, Modula-3, Guide, and Extended Ada the exceptions are 
checked, which means that the 
compiler statically checks if appropriate handlers are provided within the system. 
Accordingly, the programmer must explicitly specify every exception that can be
signaled to support the compile time checking. This approach is known to be
costly to maintain (Dooren and Steegmans, 2005) since exception updates are
cascading; if a new exception type is added to a method signature then all other
methods directly using this method must either change their signatures to
include the new exception type or handle the new exception.

@@

In Java a method can throw either checked or unchecked exceptions. Java is the only language that allows a hybrid approach on which the developer can define exceptions of both kinds to be thrown and handled in a single system. Next section details Java exception handling model.

Although exception handling mechanism is the most used technique to deal with such 
failures - been embedded in several programming languages as show above,   
the exception handling code is often generally poorly understood and the least
tested part of software systems.This behaviour may lead to the problem of uncaught exceptions~\cite{jo2004uncaught}, one of the main causes of application crashes
(e.g.~\cite{miller1997issues,Robil00,shah2010understanding, garcia2007extracting, garcia2001comparative,cabral2007exception,coelho2011unveiling}).

\subsection{Checked an Unchecked Exceptions in Java} \label{sec:extypes}

@@

Figure~\ref{fig:overviewfig} presents an overview of the mixed-methods approach
conducted to answer these questions. The mixed-methods approach is composed by the following steps:

@@

\textbf{Step1.} To obtain the Dataset to be analized, we queried GHTorrent as detailed in Section XXX. From GHTorrent we retrieved the metadata related to each Java project (e.g., 
repository's names and short descriptions), and the content of every issue and issue comment 
and stored in a relational database.

\textbf{Step 2.} Extracting Exception Information from Issues. In this step, we used the ExceptionMiner tool,
 developed in this work, to extract the exception names and exception stack traces of each
 issue and issue comment - analyzing the information of each project at a time. This initial 
set of analyzed Java projects included libraries, web-based, Android and stand-alone applications.

\textbf{Step 3.} Restricting the Dataset to Android Projects. We select a subset of all Android 
projects available on the dataset to perform a deeper analysis (which combined mined
 information extrated from issues with source-code analysis and manual inspection 
of projects resources described on next steps). This decision was driven by the
fact that most open source Anedroid applications are known for dealing
with several sources of exceptions due to multi-threaded execution and
interaction with input/output resources; moreover they share a common
well-defined platform and a limited set of libraries.
To identify Android projects, we performed a case insensitive search for 
the term \textsf{android} in the repository's names and short descriptions.  The
heuristic filtered 2.542 repositories, from which 589 apps had at least one
issue containing a stack trace. Then we performed further cleanup, inspecting the site of every Android
reporting at least one stack trace, to make sure that they represented real
mobile apps. During this clean up 106 apps were removed because they were either
example projects (i.e., toy projects) or tools to support Android development
(e.g. \textsf{selendroid}, \textsf{roboeletric} --- tools to support the testing of Android apps).
The filtered set consisted of 482 apps, from which approximately 50\% are also
hosted in Google Market Place.

To support a deeper investigation of the stack traces of Android apps, every
exception defined on a stack trace was then classified according to its type
(e.g. Error, Exception or RuntimeException) and to its signaler. To do so, the following 
steps were performed:

\textbf{Step 4.}  Bytecode-Code analysis of JVM and Android Exceptions. We developerd a bytecode analysis tool based on DesignWizard to discover all exceptions thrown by the last version of SDK version 7, and Android Platform and OS (level 19). Every exception was extracted and classified as Runtime, Checked or Error based on the analysis of the exception hierarchy. Our analysis found XXX runtime exceptions in JVM and XX checked, on Android xxx checked and YYY runtime.

\textbf{Step 5.}  Source-Code Analysis of Android Projects. 
For each one of the analyzed projects, we
downloaded the source code and using custom scripts, we extracted their package
hierarchy (recursive list of of Java packages) and their custom exception types.

\textbf{Step 6.}  Manual Inspection. XXXX of the exceptions reported on stack traces  could not be found on JMV, SO or inside the last version of the project. Based on the full name of the exception, a manual inspection was performed on the artifacts associated to the project and libraries used by the project. When even on those sources the exception was not foung [grep-code and similar sites] the exception, in order to discover the exception type. Only 3 exceptions remained undefined and were removed from this study.

OBS: THE PICTURE WILL BE UPDATED TO REFLECT THIS PROCESS DESCRIPTION.

@@

If we perform a cross-project analysis, the java.lang.NullPointerException, we could observe that such 
exception was reported on 1,671 projects (which represents 44,47\% of all projects reporting at least one stack trace).

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{tab:typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~\ref{tab:typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. After filtering all the exceptions implicitly signaled by 
JVM (due to programming mistakes) and inspecting the signaler methods of such exceptions (i.e.,  direct an indirect 
instances of java.lang.RuntimeException). We could observe that only 1 method out of 118 inspected signalers
 (i.e., 0,8\%) documented the explicitly thrown runtime exception in a Javadoc comment; and none of these methods
included the runtime exception  (reported on the stack) as part of the exception interface of the method (i.e., using 
throws clause on method signature). This result is aligned with the results of other  study conducted by from 
Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

@@

%\begin{figure}
%  \begin{center}
%    \includegraphics[scale=0.4]{stack-wrappings-hist.pdf}
%  \end{center}
%  \caption{Number of stack traces wrappings for Java and Android repositories. }
%  \label{fig:sizewrapbhists}
%\end{figure}
